Persistence in functional topology: existence of persistence diagrams and Morse inequalities

### 2022-03-28 sec/surfaces.tex
Before addressing reviewer

Having established topological conditions for the existence of persistence diagrams associated to filtrations and corresponding generalized Morse inequalities, we now illustrate how these results relate to Morse and Tompkins' work on minimal surfaces from \cite{Morse.1939}.
We will focus on the homological aspects of their approach.
For a general exposition we refer to \cite[Sections 4.3--5]{Bott.1980}, for a more detailed exposition including the analytical details see \cite[Section II.6]{Struwe.1988}, and for a historical account and an overview of subsequent results see \cite[Section 6.7]{Dierkes.2010} and \cite[Section 6.8.1]{Dierkes.2010b}.

@@

Morse and Tompkins considered the following setting introduced by Douglas.
Let $g \colon \R \to \R^n$ be a $2\pi$-periodic function representing a simple closed curve such that $g$ is differentiable with Lipschitz derivative.
Let $\widetilde{\Omega}$ be the space of continuous functions $\varphi \colon \R \to \R$ with $\varphi((t)+2\pi) = \varphi(t) + 2\pi$ for all $t$ and $\varphi(\alpha_i)=\alpha_i$ for three fixed distinct points $\alpha_i \in [0,2\pi)$.
The \emph{Douglas functional} on $\widetilde \Omega$ associated to the curve $g$ is defined as
\begin{equation*}
A_g(\varphi) = \frac{1}{16 \pi} \int_0^{2\pi} \int_0^{2\pi}  \left\| \frac{g(\varphi(\alpha)) - g(\varphi(\beta))}{\sin \frac{\alpha-\beta}{2}} \right\|_2^2 \ \mathrm{d}\alpha \ \mathrm{d}\beta.
\end{equation*}
Let $\Omega_g = \{\varphi \in \widetilde\Omega \mid A_g(\varphi) < \infty\}$, equipped with the $C^0$ metric.
Douglas proved that, for a large class of curves~$g$, the set $\Omega_g$ is non-empty and the sublevel sets of $A_g$ are compact.
Since $A_g$ is bounded below by $0$, this implies that $A_g$ attains a global minimum.
Douglas then showed that this minimizing critical point of $A_g$ corresponds to a solution of Plateau's Problem, a surface homeomorphic to a disk with boundary $g$ and minimum area.

@@

\begin{thm}[{Unstable Minimal Surface Theorem \cite[p.~472]{Morse.1939}}]
	If the space $\Omega_g$ contains two distinct solutions of Plateau's Problem contained in disjoint critical sets of the functional $A_g$, then it also contains a critical set not of minimum type.
	%, more precisely, a homotopically critical point of $A_g$ of non-minimum type.
\end{thm}
Here, a \emph{critical set} $S$ is a closed and open subspace of the set of all critical points with a given function value.
It is said to be of \emph{minimum type} if for every neighborhood $N$ of $S$, the function value on the complement $N \setminus S$ exceeds the function value on $S$. 
%it corresponds to a homotopically critical point of $A_g$ of non-minimum type?
In addition to the general theorem above, Morse and Tompkins also give an explicit example of a curve $g$ satisfying the assumptions, thus proving the existence of an unstable minimal surface.

While more efficient and more general proofs for the existence of an unstable minimal surface (with respect to more natural topologies than $C^0$) have subsequently been established \cite{Struwe.1988,Dierkes.2010}, the original approach of Morse and Tompkins is notable for its connection to persistent homology.

@@

Morse and Tompkins' proof of the existence of unstable minimal surfaces is revisited in \cite{Morse.1940} using an argument that relies on Morse inequalities for cap numbers associated to the persistent \v{C}ech homology of the sublevel set filtration, as also considered in \cref{s:inequalities}.

We have seen that these inequalities require q-tameness, so in order to apply them and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, in order to obtain \mbox{q-tameness}, Morse assumed slightly varying forms of local connectivity on the resulting sublevel set filtrations.
In particular, in their applications to minimal surface theory, Morse and Tompkins used the following condition:
\begin{displaycquote}[p.~431]{Morse.1940}
	Let $p$ be a point of $M$ at which $F(p)=c$.
	The space $M$ is said to be \emph{locally $F$-connected} of order $r$ at~$p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ and on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Using similar language to the one used in \cref{s:connectivity}, the property of local $F$-connectedness of all orders is equivalent to the following notion applicable to general topological spaces.

@@

Specifically, \v{C}ech homology is infinite dimensional in degree $d$ for the $d$-dimensional Hawaiian earring. 
To see that his is the case, one can use the fact that \v{C}ech homology commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6 and X.3.1]{Eilenberg.1952} as follows.
Define
\begin{align*}
\HE_k &=
\left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle| \ \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\ &\, \cup
\bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle |\ \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \F) = \lim_{k} \CH_{d}(\HE_{k}; \F)$, where $\CH$ denotes \v{C}ech homology.
Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
\begin{equation*}
\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k}; \F)=\lim\left(\dots\to \prod_{n=1}^2\F\to \prod_{n=1}^1\F\to \prod_{n=1}^0\F\right)=\prod_{n\in\N}\F,
\end{equation*}
which is infinite-dimensional over $\F$.

%We will consider the \emph{$d$-dimensional Hawaiian earring}
%\begin{equation*}
%\HE = \bigcup_{n \in \N} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle | \ \left( x_0 - \frac{1}{n} \right)^2 \!\! + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
%\end{equation*}
%which is a compact subspace of $\R^{d+1}$.
%
%\begin{figure}[t]
%	\centering
%	\begin{tikzpicture}[scale = 60]
%	\draw[thick] (-.1,-.001) rectangle (.1,.05);
%	\clip (-.1,0) rectangle (.1,.05);
%	\foreach \i in {1,...,100}{
%		\draw[line width=0.4/\i^0.25 pt] (0, 1/\i^2) circle (1/\i^2);
%	}
%	\end{tikzpicture}
%	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
%\end{figure}
%
%\begin{thm} \label{t:counterexample}
%	The function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a compact and weakly $\LC$ sublevel set filtration that is not q-tame with respect to $\H$ if $\H_{n}(\HE)$ is infinite dimensional for some $n$.
%\end{thm}
%
%\begin{proof}
%	To verify that $f$ has compact sublevel sets we notice that all sublevel sets are either the empty set, the singleton containing the origin, or $\HE$ itself, all compact Hausdorff spaces.
%
%	In order to verify that the sublevel set filtration of $f$ is weakly $\LC$, we
%	consider $x \in \HE$, $V$ a neighborhood of $x$ in $\HE$ and $t > f(x)$. We need to find a neighborhood $U \subseteq V$ of $x$ and $s \in (f(x), t)$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to a constant map.
%
%	If $x$ is the origin, we have $f(x) = 0$ and choose $s \in (0, \min\{t, 1\})$.
%	Then $f_{\leq s} = \{x\}$, so with $U = V$ the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is the inclusion of $\{x\}$ into $f_{\leq t} \cap V$, which is a constant map, so the weak $\LC$ condition is trivially satisfied.
%
%	For $x$ not the origin we have $f(x) = 1$ and choose $s \in (1,t)$ arbitrarily, so that $f_{\leq s} = f_{\leq t} = \HE$.
%	Note that since $x$ is not the origin, there is a unique $d$-sphere in $\HE$ that contains $x$.
%	Clearly, we may choose $\delta > 0$ so small that $B_{\delta}(x) = \{y \in \R^{d+1} \mid \Vert x - y \Vert < \delta\} \cap \HE$ is a topological ball contained in this sphere and contained in $V$.
%	The ball $B_\delta(x)$ can be contracted to $\{x\}$ in $V$, so choosing $U = B_{\delta}(x)$, we obtain that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to the constant map with value $x$.
%
%	It remains to be shown that $f_{\leq \bullet}$ is not q-tame for $\H$.
%	This follows directly from our assumption that $\H_{n}(\HE)$ is not finite dimensional for some $n$, as $f_{\leq t}$ is constant with value $\HE$ for $t \geq 1$.
%\end{proof}

As proven in \cite{Barratt.1962}, the singular homology of the Hawaiian earrings is also infinite dimensional. In summary, we have the following.

@@

The gap we have thus highlighted in the argument of Morse and Tompkins, that the sublevel set filtration of $A_g$ is not necessarily q-tame, can be fixed by applying \cref{t:local connectedness implies q-tameness}.
This is because the proof given in \cite[p.464]{Morse.1939} for the local connectivity of $(\Omega_g, A_g)$ actually establishes a stronger property described next.

@@

Clearly, the filtration being $\LC$ implies that the filtration is also $\HLC$ for \v{C}ech homology with field coefficients, which is the homology theory Morse and Tompkins use.
Therefore, from \cref{t:local connectedness implies q-tameness} we can conclude that $(\Omega_g, A_g)$ is indeed \mbox{q-tame} as needed.
This implies that the generalized Morse inequalities hold, and hence so does the Unstable Minimal Surfaces Theorem.

>>>>

Having established topological conditions for the existence of persistence diagrams associated to filtrations and corresponding generalized Morse inequalities, we now describe how these results relate to Morse's general theory of functional topology \cite{Morse.1937, Morse.1938, Morse.1940}, and to its application in Morse and Tompkins' work on unstable minimal surfaces from \cite{Morse.1939}.
We will focus on the homological aspects of this approach referring to \cite[Sections 4.3--5]{Bott.1980} for a general exposition.
For a more thorough presentation of the unstable minimal surface problem, including the analytical details, please consult \cite[Section II.6]{Struwe.1988}, and for a historical account and an overview of subsequent results see \cite[Section 6]{Dierkes.2010} and \cite[Section 6.8.1]{Dierkes.2010b}.

@@

Morse and Tompkins considered the following setting introduced by Douglas.
Let $g \colon \R \to \R^n$ be a $2\pi$-periodic function representing a simple closed curve such that $g$ is differentiable with Lipschitz derivative.
Let $\widetilde{\Omega}$ be the space of continuous functions $\varphi \colon \R \to \R$ with $\varphi((t)+2\pi) = \varphi(t) + 2\pi$ for all $t$ and $\varphi(\alpha_i)=\alpha_i$ for three fixed distinct points $\alpha_i \in [0,2\pi)$.
The \emph{Douglas functional} on $\widetilde \Omega$ associated to the curve $g$ is defined as
\begin{equation*}
A_g(\varphi) = \frac{1}{16 \pi} \int_0^{2\pi} \int_0^{2\pi}  \left\| \frac{g(\varphi(\alpha)) - g(\varphi(\beta))}{\sin \frac{\alpha-\beta}{2}} \right\|_2^2 \ \mathrm{d}\alpha \ \mathrm{d}\beta.
\end{equation*}
It coincides with the Dirichlet energy of the unique harmonic extension of the reparametrized curve $g \circ \phi$ to a parametrized surface.
The Dirichlet energy is an upper bound for the area, with equality if the parametrization is conformal.
Let $\Omega_g = \{\varphi \in \widetilde\Omega \mid A_g(\varphi) < \infty\}$, equipped with the $C^0$ metric.
Douglas proved that, for a large class of curves~$g$, the set $\Omega_g$ is non-empty and the sublevel sets of $A_g$ are compact.
Since $A_g$ is bounded below by $0$, this implies that $A_g$ attains a global minimum.
The corresponding surface is then a solution of \emph{Plateau's Problem}, which asks for a surface homeomorphic to a disk with boundary $g$ and minimum area.

@@

\begin{thm}[{Unstable Minimal Surface Theorem \cite[p.~472]{Morse.1939}}]
	If the space $\Omega_g$ contains two distinct solutions of Plateau's Problem contained in disjoint critical sets of the functional $A_g$, then it also contains a critical set not of minimum type.
	%, more precisely, a homotopically critical point of $A_g$ of non-minimum type.
\end{thm}
Here, a \emph{critical set} $S$ is a closed and open subspace of the set of all critical points with a given function value.
It is said to be of \emph{minimum type} if for every neighborhood $N$ of $S$, the function value on the complement $N \setminus S$ exceeds the function value on $S$. 
%it corresponds to a homotopically critical point of $A_g$ of non-minimum type?
In addition to the general theorem above, Morse and Tompkins also give an explicit example of a curve $g$ satisfying the assumptions, thus proving the existence of an unstable minimal surface in the above sense.

%sci-hub.se/10.1215/S0012-7094-41-00828-1
%
%A point q of M will be termed non-minimizing if in every neighborhood of q there is a point p such thatF(p)<F(q). A homotopic critical set which is not a minimizing set contains at least one non-minimizing point of F.

While more efficient and more general proofs for the existence of an unstable minimal surface (with respect to more natural topologies than $C^0$) have subsequently been established \cite{Struwe.1988,Dierkes.2010}, including less restrictive assumptions on the boundary curve, the original approach of Morse and Tompkins is notable for its close connection to persistent homology.

@@

Parts of the framework of functional topology that Morse and Tompkins use is developed by Morse in \cite{Morse.1940} in a very general setting. 
In particular, he proves inequalities for cap numbers associated to the persistent \v{C}ech homology of the sublevel set filtration, as also considered in \cref{s:inequalities}.
From these generalized Morse inequalities, the existence of an unstable minimal surface can easily be deduced in the presence of two distinct critical set of minimum type.

We have seen that these inequalities require q-tameness, so in order to apply them in the minimal surface setting % and deduce their theorem,
one needs to prove the \mbox{q-tameness} of the sublevel set filtration of the Douglas functional $A_g$.
Throughout his work on functional topology, in order to obtain \mbox{q-tameness}, Morse assumed slightly varying forms of local connectivity on the resulting sublevel set filtrations.
In particular, Morse and Tompkins used the following condition in their applications to minimal surface theory (see also \cite[p.~ 25]{Morse.1938} and \cite[p.~464]{Morse.1939}, but note that the definitions given there contain typographical errors):
\begin{displaycquote}[p.~431]{Morse.1940}
%\cite[p.~ 25]{Morse.1938}
	Let $p$ be a point of $M$ at which $F(p)=c$.
	The space $M$ is said to be \emph{locally $F$-connected} of order $r$ at~$p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ and on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Using similar language to the one used in \cref{s:connectivity}, the property of local $F$-connectedness of all orders is equivalent to the following notion applicable to general topological spaces.

@@

Specifically, using the the fact that \v{C}ech homology of compact Hausdorff spaces commutes with inverse limits, it is straightforward to verify that the \v{C}ech homology in degree $d$ for the $d$-dimensional Hawaiian earring is isomorphic to $\prod_{n\in\N}\F$, which is infinite dimensional over $\F$. 
%To see that his is the case, one can use the fact that \v{C}ech homology commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6 and X.3.1]{Eilenberg.1952} as follows.
%Define
%\begin{align*}
%\HE_k &=
%\left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle| \ \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\ &\, \cup
%\bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle |\ \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
%\end{align*}
%i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
%We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \F) = \lim_{k} \CH_{d}(\HE_{k}; \F)$, where $\CH$ denotes \v{C}ech homology.
%Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
%\begin{equation*}
%\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k}; \F)=\lim\left(\dots\to \prod_{n=1}^2\F\to \prod_{n=1}^1\F\to \prod_{n=1}^0\F\right)=\prod_{n\in\N}\F,
%\end{equation*}
%which is infinite-dimensional over $\F$.
%We will consider the \emph{$d$-dimensional Hawaiian earring}
%\begin{equation*}
%\HE = \bigcup_{n \in \N} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle | \ \left( x_0 - \frac{1}{n} \right)^2 \!\! + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
%\end{equation*}
%which is a compact subspace of $\R^{d+1}$.
%
%\begin{figure}[t]
%	\centering
%	\begin{tikzpicture}[scale = 60]
%	\draw[thick] (-.1,-.001) rectangle (.1,.05);
%	\clip (-.1,0) rectangle (.1,.05);
%	\foreach \i in {1,...,100}{
%		\draw[line width=0.4/\i^0.25 pt] (0, 1/\i^2) circle (1/\i^2);
%	}
%	\end{tikzpicture}
%	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
%\end{figure}
%
%\begin{thm} \label{t:counterexample}
%	The function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a compact and weakly $\LC$ sublevel set filtration that is not q-tame with respect to $\H$ if $\H_{n}(\HE)$ is infinite dimensional for some $n$.
%\end{thm}
%
%\begin{proof}
%	To verify that $f$ has compact sublevel sets we notice that all sublevel sets are either the empty set, the singleton containing the origin, or $\HE$ itself, all compact Hausdorff spaces.
%
%	In order to verify that the sublevel set filtration of $f$ is weakly $\LC$, we
%	consider $x \in \HE$, $V$ a neighborhood of $x$ in $\HE$ and $t > f(x)$. We need to find a neighborhood $U \subseteq V$ of $x$ and $s \in (f(x), t)$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to a constant map.
%
%	If $x$ is the origin, we have $f(x) = 0$ and choose $s \in (0, \min\{t, 1\})$.
%	Then $f_{\leq s} = \{x\}$, so with $U = V$ the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is the inclusion of $\{x\}$ into $f_{\leq t} \cap V$, which is a constant map, so the weak $\LC$ condition is trivially satisfied.
%
%	For $x$ not the origin we have $f(x) = 1$ and choose $s \in (1,t)$ arbitrarily, so that $f_{\leq s} = f_{\leq t} = \HE$.
%	Note that since $x$ is not the origin, there is a unique $d$-sphere in $\HE$ that contains $x$.
%	Clearly, we may choose $\delta > 0$ so small that $B_{\delta}(x) = \{y \in \R^{d+1} \mid \Vert x - y \Vert < \delta\} \cap \HE$ is a topological ball contained in this sphere and contained in $V$.
%	The ball $B_\delta(x)$ can be contracted to $\{x\}$ in $V$, so choosing $U = B_{\delta}(x)$, we obtain that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to the constant map with value $x$.
%
%	It remains to be shown that $f_{\leq \bullet}$ is not q-tame for $\H$.
%	This follows directly from our assumption that $\H_{n}(\HE)$ is not finite dimensional for some $n$, as $f_{\leq t}$ is constant with value $\HE$ for $t \geq 1$.
%\end{proof}
Moreover, the singular homology of the $d$-dimensional Hawaiian earring is also infinite dimensional, as proven in \cite{Barratt.1962}. In summary, we have the following.

@@

The gap we have thus highlighted in the argument of Morse and Tompkins, that the sublevel set filtration of $A_g$ is not necessarily q-tame, can be fixed by applying \cref{t:local connectedness implies q-tameness}.
This is because the proof given in \cite[p.464]{Morse.1939} for the local connectivity of the sublevel set filtration induced by $A_g$ can actually be seen to establish a stronger property described next.

@@

Clearly, the filtration being $\LC$ implies that the filtration is also $\HLC$ for \v{C}ech homology with field coefficients, which is the homology theory Morse and Tompkins use.
Therefore, from \cref{t:local connectedness implies q-tameness} we can conclude that the sublevel set filtration of $A_g$ is indeed \mbox{q-tame} as needed.
This implies that the generalized Morse inequalities hold, and hence so does the Unstable Minimal Surfaces Theorem.
